---
title: "Complete Guide to surveytoolbox"
subtitle: "A comprehensive walkthrough of survey analysis functions"
author: "surveytoolbox package"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Complete Guide to surveytoolbox}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, message=FALSE}
library(surveytoolbox)
library(dplyr)
library(tibble)
library(haven)
library(purrr)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)
```

# Introduction

The `surveytoolbox` package provides a comprehensive set of functions designed specifically for survey data analysis in R. This vignette demonstrates the key functions organized by their main use cases:

1. **Data Labeling and Management**
2. **Scale and Variable Transformations** 
3. **Survey-Specific Calculations**
4. **Data Cleaning and Preparation**
5. **Utility and Helper Functions**

Let's start by creating some sample survey data to demonstrate these functions.

# Sample Survey Dataset

```{r create_sample_data}
# Create a sample survey dataset
n <- 500

survey_data <- tibble(
  respondent_id = 1:n,
  age = sample(18:80, n, replace = TRUE),
  gender = sample(1:3, n, replace = TRUE, prob = c(0.45, 0.45, 0.1)),
  satisfaction = sample(1:7, n, replace = TRUE, prob = c(0.05, 0.1, 0.15, 0.2, 0.25, 0.15, 0.1)),
  nps_score = sample(0:10, n, replace = TRUE),
  income = sample(1:5, n, replace = TRUE),
  usage_freq = sample(1:5, n, replace = TRUE),
  recommend = sample(1:2, n, replace = TRUE, prob = c(0.3, 0.7)),
  brand_awareness_1 = sample(0:1, n, replace = TRUE, prob = c(0.6, 0.4)),
  brand_awareness_2 = sample(0:1, n, replace = TRUE, prob = c(0.7, 0.3)),
  brand_awareness_3 = sample(0:1, n, replace = TRUE, prob = c(0.8, 0.2)),
  open_text = paste("Response", 1:n),
  completion_time = runif(n, 5, 45)
)

head(survey_data)
```

# 1. Data Labeling and Management

## Setting Variable Labels with `set_varl()`

Variable labels provide descriptive names for your variables, essential for survey data documentation.

```{r variable_labels}
# Add variable labels to our survey data
survey_labeled <- survey_data %>%
  mutate(
    respondent_id = set_varl(respondent_id, "Unique Respondent Identifier"),
    age = set_varl(age, "Respondent Age in Years"),
    gender = set_varl(gender, "Q1. What is your gender?"),
    satisfaction = set_varl(satisfaction, "Q2. Overall satisfaction with our service"),
    nps_score = set_varl(nps_score, "Q3. Net Promoter Score (0-10)"),
    income = set_varl(income, "Q4. Annual household income bracket"),
    usage_freq = set_varl(usage_freq, "Q5. How often do you use our service?"),
    recommend = set_varl(recommend, "Q6. Would you recommend us to others?")
  )

# Check variable labels
attr(survey_labeled$satisfaction, "label")
attr(survey_labeled$nps_score, "label")
```

## Setting Value Labels with `set_vall()`

Value labels map numeric codes to meaningful text descriptions.

```{r value_labels}
# Add value labels
survey_labeled <- survey_labeled %>%
  mutate(
    gender = set_vall(gender, c("Male" = 1, "Female" = 2, "Other/Prefer not to say" = 3)),
    satisfaction = set_vall(satisfaction, c(
      "Extremely dissatisfied" = 1,
      "Very dissatisfied" = 2,
      "Somewhat dissatisfied" = 3,
      "Neither satisfied nor dissatisfied" = 4,
      "Somewhat satisfied" = 5,
      "Very satisfied" = 6,
      "Extremely satisfied" = 7
    )),
    income = set_vall(income, c(
      "Under $25k" = 1,
      "$25k-$50k" = 2,
      "$50k-$75k" = 3,
      "$75k-$100k" = 4,
      "Over $100k" = 5
    )),
    usage_freq = set_vall(usage_freq, c(
      "Never" = 1,
      "Rarely" = 2,
      "Sometimes" = 3,
      "Often" = 4,
      "Very often" = 5
    )),
    recommend = set_vall(recommend, c("No" = 1, "Yes" = 2))
  )

# Check the labelled structure
str(survey_labeled$gender)
```

## Creating Variable Label Tables with `varl_tb()`

This function creates a tidy data frame showing all variable labels in your dataset.

```{r variable_table}
# Create a variable label table
var_labels <- varl_tb(survey_labeled)
print(var_labels)
```

## Creating Data Dictionaries with `data_dict()`

Generate comprehensive data dictionaries showing variable information, labels, and values.

```{r data_dictionary}
# Create a data dictionary for key survey variables
survey_labeled %>%
  select(gender, satisfaction, income, usage_freq, recommend) %>%
  data_dict()
```

# 2. Scale and Variable Transformations

## Creating Binary Variables with `box_it()`

The `box_it()` function converts ordinal scales into binary "top box" or "bottom box" variables.

```{r box_it_examples}
# Create top-2-box satisfaction score
satisfaction_t2b <- box_it(
  survey_data$satisfaction, 
  which = "top", 
  number = 2,
  var_label = "Satisfaction Top-2-Box (Very/Extremely Satisfied)"
)

# Create bottom-3-box (dissatisfied responses)
satisfaction_b3b <- box_it(
  survey_data$satisfaction,
  which = "bottom",
  number = 3,
  var_label = "Satisfaction Bottom-3-Box (Dissatisfied)"
)

# Check results
table(satisfaction_t2b, survey_data$satisfaction)
```

## Reversing Likert Scales with `likert_reverse()`

Sometimes you need to reverse-code survey items for analysis.

```{r likert_reverse}
# Reverse a satisfaction scale (7-point to 1-point becomes 1-point to 7-point)
satisfaction_reversed <- likert_reverse(survey_data$satisfaction, top = 7, bottom = 1)

# Compare original and reversed
comparison <- data.frame(
  original = survey_data$satisfaction[1:10],
  reversed = satisfaction_reversed[1:10]
)
print(comparison)
```

## Converting Likert Scales with `likert_convert()`

Convert between different scale ranges (e.g., 7-point to 10-point scale).

```{r likert_convert}
# Convert 7-point satisfaction to 10-point scale
satisfaction_10pt <- likert_convert(
  survey_data$satisfaction, 
  top.x = 7, bot.x = 1,    # Original scale
  top.y = 10, bot.y = 1    # New scale
)

# Show conversion
conversion_example <- data.frame(
  original_7pt = survey_data$satisfaction[1:10],
  converted_10pt = satisfaction_10pt[1:10]
)
print(conversion_example)
```

## Max-Min Scaling with `maxmin()`

Normalize variables to a 0-1 scale using max-min normalization.

```{r maxmin_scaling}
# Scale age to 0-1 range
age_scaled <- maxmin(survey_data$age)

# Show original vs scaled
scaling_example <- data.frame(
  original_age = survey_data$age[1:10],
  scaled_age = round(age_scaled[1:10], 3)
)
print(scaling_example)

# Verify scaling worked correctly
cat("Original age range:", min(survey_data$age), "to", max(survey_data$age), "\n")
cat("Scaled age range:", round(min(age_scaled), 3), "to", round(max(age_scaled), 3), "\n")
```

# 3. Survey-Specific Calculations

## Net Promoter Score (NPS) with `as_nps()`

Convert 0-10 scores to NPS categories and calculate NPS.

```{r nps_analysis}
# Convert NPS scores to categories
nps_categorized <- as_nps(survey_data$nps_score)

# Check the labelled structure
table(nps_categorized)

# Calculate overall NPS (should be around 0 for our random data)
overall_nps <- mean(nps_categorized, na.rm = TRUE)
cat("Overall NPS:", round(overall_nps, 1), "\n")

# NPS by gender
nps_by_gender <- survey_labeled %>%
  mutate(nps_cat = as_nps(nps_score)) %>%
  group_by(gender) %>%
  summarise(
    nps = round(mean(nps_cat, na.rm = TRUE), 1),
    n = n(),
    .groups = 'drop'
  )
print(nps_by_gender)
```

## Alternative NPS Categories with `as_nps_cat()`

Create custom NPS categories with different numeric codes.

```{r nps_custom}
# Create NPS categories with custom coding
nps_custom <- as_nps_cat(
  survey_data$nps_score,
  det = 1,    # Detractors coded as 1
  pas = 2,    # Passives coded as 2  
  pro = 3     # Promoters coded as 3
)

table(nps_custom)
```

## CAGR Calculations with `CAGR()`

Calculate Compound Annual Growth Rate - useful for longitudinal survey metrics.

```{r cagr_example}
# Example: Customer satisfaction improved from 60% to 75% over 3 years
satisfaction_cagr <- CAGR(
  value_begin = 60, 
  value_end = 75, 
  n_periods = 3
)

cat("Annual satisfaction improvement rate:", round(satisfaction_cagr * 100, 2), "%\n")

# Example: NPS growth from 20 to 45 over 2 years
nps_cagr <- CAGR(20, 45, 2)
cat("Annual NPS growth rate:", round(nps_cagr * 100, 2), "%\n")
```

## Percentage Formatting with `as_percent()`

Convert numeric values to formatted percentage strings.

```{r as_percent}
# Calculate satisfaction rates
satisfaction_rates <- survey_labeled %>%
  group_by(gender) %>%
  summarise(
    satisfied_rate = mean(satisfaction >= 5, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    satisfied_percent = as_percent(satisfied_rate, rounding = 1)
  )

print(satisfaction_rates)

# Example with different rounding
example_values <- c(0.1234, 0.5678, 0.9999)
cat("No rounding:", as_percent(example_values), "\n")
cat("1 decimal:", as_percent(example_values, rounding = 1), "\n")
cat("2 decimals:", as_percent(example_values, rounding = 2), "\n")
```

# 4. Data Cleaning and Preparation

## String Cleaning with `clean_strings()`

Clean variable names and text for analysis and visualization.

```{r clean_strings}
# Example survey question texts
messy_questions <- c(
  "Q1. What is your gender?",
  "Q23. Brand Awareness - Company A",
  "Q45. Overall satisfaction (%)",
  "'Net Promoter Score'",
  "  Usage frequency  "
)

# Clean them for use as variable names
clean_names <- clean_strings(messy_questions)
print(data.frame(original = messy_questions, cleaned = clean_names))

# Handle duplicates
duplicate_questions <- c("Satisfaction", "Satisfaction", "Satisfaction")
clean_dupes <- clean_strings(duplicate_questions, treat_dups = TRUE)
print(clean_dupes)
```

## Text Wrapping with `wrap_text()`

Wrap long text for better visualization in charts and tables.

```{r wrap_text}
# Wrap long question text for plotting
long_questions <- c(
  "Overall, how satisfied are you with our customer service experience?",
  "Would you recommend our product to your friends and colleagues?",
  "How likely are you to purchase from us again in the future?"
)

wrapped_questions <- wrap_text(long_questions, threshold = 30)
cat("Original:\n", long_questions[1], "\n\n")
cat("Wrapped:\n", wrapped_questions[1], "\n")
```

## Enhanced `any()` Function with `any_x()`

A more nuanced version of `any()` that handles all-NA cases appropriately.

```{r any_x_demo}
# Example: Brand awareness across multiple brands
brand_data <- tibble(
  respondent = 1:6,
  brand_a = c(1, 0, 1, NA, NA, 0),
  brand_b = c(0, 1, 0, NA, NA, 1), 
  brand_c = c(1, 1, 0, NA, NA, 0)
)

# Check if respondent is aware of ANY brand
brand_data$any_awareness <- apply_row(brand_data[2:4], everything(), function(x) any_x(x, 1))

print(brand_data)

# Compare with regular any() function
brand_data$any_regular <- apply_row(brand_data[2:4], everything(), function(x) any(x == 1, na.rm = TRUE))

# Show the difference for all-NA cases
comparison <- brand_data %>% 
  filter(respondent %in% c(4, 5)) %>%
  select(respondent, any_awareness, any_regular)
print(comparison)
```

## Row-wise Operations with `apply_row()`

Apply functions across rows of selected columns.

```{r apply_row_examples}
# Calculate row-wise statistics
survey_stats <- survey_data %>%
  select(brand_awareness_1, brand_awareness_2, brand_awareness_3) %>%
  mutate(
    total_awareness = apply_row(., everything(), sum, na.rm = TRUE),
    max_awareness = apply_row(., everything(), max, na.rm = TRUE),
    any_awareness = apply_row(., everything(), function(x) any_x(x, 1))
  )

head(survey_stats)

# More complex example: satisfaction across multiple dimensions
# Let's create some multi-dimensional satisfaction data
satisfaction_dims <- tibble(
  service_sat = sample(1:7, n, replace = TRUE),
  product_sat = sample(1:7, n, replace = TRUE),
  value_sat = sample(1:7, n, replace = TRUE)
)

satisfaction_summary <- satisfaction_dims %>%
  mutate(
    mean_satisfaction = apply_row(., everything(), mean, na.rm = TRUE),
    top_box_count = apply_row(., everything(), function(x) sum(x >= 6, na.rm = TRUE)),
    all_satisfied = apply_row(., everything(), function(x) all(x >= 5, na.rm = TRUE))
  )

head(satisfaction_summary)
```

# 5. Utility and Helper Functions

## Lookup Tables with `look_up()`

Map values using lookup tables - useful for recoding survey responses.

```{r lookup_examples}
# Create a lookup table for region codes
region_lookup <- data.frame(
  code = 1:5,
  region = c("North", "South", "East", "West", "Central"),
  stringsAsFactors = FALSE
)

# Generate some region codes
region_codes <- sample(1:5, 20, replace = TRUE)

# Map to region names
region_names <- look_up(region_codes, region_lookup, index = "code", column = "region")

# Show mapping
mapping_example <- data.frame(
  code = region_codes[1:10],
  region = region_names[1:10]
)
print(mapping_example)
```

## Creating Named Lists with `create_named_list()`

Create named lists for value labels and other uses.

```{r named_lists}
# Create a named list for survey responses
response_options <- c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree")
response_codes <- 1:5

# Create named list
response_list <- create_named_list(response_options, response_codes)
print(response_list)
```

## Squish Function for Data Consistency

The `squish()` function ensures data consistency by returning single values when all values in a group are identical.

```{r squish_example}
# Example: Demographics should be consistent within household
household_data <- tibble(
  household_id = c(1, 1, 1, 2, 2, 3, 3, 3),
  person_id = 1:8,
  household_income = c(75000, 75000, 75000, 50000, 50000, 100000, 100000, 100000),
  zip_code = c("12345", "12345", "12345", "67890", "67890", "54321", "54321", "54321")
)

# Aggregate to household level
household_summary <- household_data %>%
  group_by(household_id) %>%
  summarise(
    income = squish(household_income),
    zip = squish(zip_code),
    household_size = n(),
    .groups = 'drop'
  )

print(household_summary)
```

# 6. Practical Workflow Example

Let's put it all together in a typical survey analysis workflow:

```{r complete_workflow}
# 1. Start with raw survey data
survey_analysis <- survey_data %>%
  
  # 2. Add variable labels
  mutate(
    satisfaction = set_varl(satisfaction, "Q2. Overall Satisfaction (7-point scale)"),
    nps_score = set_varl(nps_score, "Q3. Net Promoter Score"),
    gender = set_varl(gender, "Q1. Gender")
  ) %>%
  
  # 3. Add value labels
  mutate(
    gender = set_vall(gender, c("Male" = 1, "Female" = 2, "Other" = 3)),
    satisfaction = set_vall(satisfaction, c(
      "Extremely dissatisfied" = 1, "Very dissatisfied" = 2, 
      "Somewhat dissatisfied" = 3, "Neutral" = 4,
      "Somewhat satisfied" = 5, "Very satisfied" = 6, 
      "Extremely satisfied" = 7
    ))
  ) %>%
  
  # 4. Create derived variables
  mutate(
    # Top-2-box satisfaction
    satisfaction_t2b = box_it(satisfaction, "top", 2, var_label = "Satisfaction T2B"),
    
    # NPS categories  
    nps_category = as_nps(nps_score),
    
    # Scaled age
    age_scaled = maxmin(age),
    
    # Brand awareness summary
    total_brand_awareness = apply_row(
      select(., starts_with("brand_awareness")), 
      everything(), 
      sum, na.rm = TRUE
    )
  )

# 5. Create summary analysis
summary_results <- survey_analysis %>%
  group_by(gender) %>%
  summarise(
    n = n(),
    avg_age = round(mean(age, na.rm = TRUE), 1),
    satisfaction_rate = as_percent(mean(satisfaction_t2b == 1, na.rm = TRUE), 1),
    avg_nps = round(mean(nps_category, na.rm = TRUE), 1),
    avg_brand_awareness = round(mean(total_brand_awareness, na.rm = TRUE), 1),
    .groups = 'drop'
  )

print(summary_results)

# 6. Create a data dictionary for the final dataset
final_dict <- survey_analysis %>%
  select(gender, satisfaction, satisfaction_t2b, nps_score, nps_category) %>%
  data_dict()

print(final_dict)
```

# 7. Advanced Examples

## Working with Missing Data

```{r missing_data}
# Create data with missing values
survey_missing <- survey_data %>%
  mutate(
    # Introduce some missing values
    satisfaction = ifelse(runif(n()) < 0.1, NA, satisfaction),
    nps_score = ifelse(runif(n()) < 0.15, NA, nps_score)
  )

# Use any_x to handle missing data appropriately
brand_awareness_any <- survey_missing %>%
  select(starts_with("brand_awareness")) %>%
  mutate(
    # This handles all-NA rows correctly
    any_brand_aware = apply_row(., everything(), function(x) any_x(x, 1))
  )

# Check cases where all brand awareness questions are NA
all_na_cases <- which(is.na(brand_awareness_any$any_brand_aware))
cat("Cases with all-NA brand awareness:", length(all_na_cases), "\n")
```

## Scale Conversion Workflow

```{r scale_conversion}
# Convert multiple scales for comparison
scale_comparison <- survey_data %>%
  mutate(
    # Original 7-point satisfaction
    satisfaction_7pt = satisfaction,
    
    # Convert to 5-point scale
    satisfaction_5pt = likert_convert(satisfaction, 7, 1, 5, 1),
    
    # Convert to 10-point scale  
    satisfaction_10pt = likert_convert(satisfaction, 7, 1, 10, 1),
    
    # Create binary satisfied/not satisfied
    satisfaction_binary = box_it(satisfaction, "top", 3, var_label = "Satisfied (Binary)")
  ) %>%
  select(respondent_id, starts_with("satisfaction")) %>%
  slice(1:10)

print(scale_comparison)
```

# Conclusion

The `surveytoolbox` package provides a comprehensive set of tools for survey data analysis in R. Key benefits include:

- **Standardized labeling**: Consistent variable and value labeling following survey research best practices
- **Scale transformations**: Easy conversion between different scale types and ranges  
- **Survey-specific metrics**: Built-in support for NPS, top-box analysis, and other common survey metrics
- **Data cleaning**: Robust tools for cleaning and preparing survey data
- **Missing data handling**: Functions that appropriately handle missing survey responses

These functions work seamlessly with the tidyverse ecosystem and can be easily integrated into existing survey analysis workflows. The package is particularly valuable for analysts working with SPSS-style labeled data and those conducting regular survey research.

For more information and updates, visit the [package repository](https://github.com/martinctc/surveytoolbox).

---

```{r session_info}
sessionInfo()
```
